---
title: "R Code"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r, cache=TRUE}
library(leaps)
library(glmnet)
library(tree)
library(MASS)
library(randomForest)
library(gbm)
library(ISLR2)
library(pls)
library(mgcv)
library(ggplot2)
library(caret) 
library(splines)
```


# Load data from CSV file
```{r, cache=TRUE}
df = read.csv("wine-quality-white-and-red.csv")
head(df)
sum(is.na(df))
```

# Split data into 2 separate white wine and red wine datasets
```{r, cache=TRUE}
df_white = subset(df, type=="white")
df_white = df_white[,-1] # remove type column
head(df_white)
df_red = subset(df, type=="red")
df_red = df_red[,-1] # remove type column
head(df_red)
```

# Exploratory Data Analysis
```{r, cache=TRUE}
# counts
print("Red Wine Sample Size: ")
length(df_red$chlorides)
print("White Wine Sample Size: ")
length(df_white$chlorides)

# counts of each quality value
table(df_white$quality)
table(df_red$quality)

# Define predictor names and datasets
predictors <- c("alcohol", "volatile.acidity", "citric.acid", "sulphates", 
                "fixed.acidity", "residual.sugar", "chlorides", 
                "free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH")
# puts the separate white and red wine datasets into a list 
datasets <- list(df_red = df_red, df_white = df_white)
```

```{r, cache=TRUE}
# Distributions
for (data_name in names(datasets)) {
  data <- datasets[[data_name]]
  par(mfrow = c(2, 3))
    for (pred in predictors) {
      hist(data[[pred]], main = paste(data_name, pred), xlab = pred)
    }
}
```

```{r, cache=TRUE}
# correlation matrix
for (data_name in names(datasets)) {
  data <- datasets[[data_name]]
  print(data_name)
  cor_matrix <- cor(data[, predictors])
  print(cor_matrix)
}
```

```{r, cache=TRUE}
# find outliers with boxplots
for (data_name in names(datasets)) {
  data <- datasets[[data_name]]
  par(mfrow = c(1, 3))  # Arrange plots
    for (pred in predictors) {
      boxplot(data[[pred]], main = paste(data_name, pred), xlab = pred)
    }
}
```

## Logistic Regression
```{r, cache=TRUE}
median(df_white$quality)
median(df_red$quality)
# median is 6, but cutoff is uneven; I use five because this cutoff is slightly 
# less uneven. This is true for both white and red.
# sum(white$quality<=6)-sum(white$quality>6) = 2788
# sum(white$quality>5)-sum(white$quality<=5) = 1618
# sum(red$quality<=6)-sum(red$quality>6) = 1165
# sum(red$quality>5)-sum(red$quality<=5) = 111

dftmp_white = df_white
dftmp_red = df_red
# creating a new column for response
dftmp_white$highquality <- as.integer(dftmp_white$quality>5)
dftmp_red$highquality <- as.integer(dftmp_red$quality>5)

datasetstmp <- list(dftmp_white = dftmp_white, dftmp_red = dftmp_red)
results = list()
results_wrongs = list()
for (data_name in names(datasetstmp)) {
  df_current <- datasetstmp[[data_name]]
  set.seed(5)
  errorrate = numeric(10)
  wrongs <- c() #vector of true qualities of mispredicted low/high wines
  for (i in 1:10){
    trainind <- sample(1:nrow(df_current), nrow(df_current)/2)
    
    df_currenttrain <- df_current[trainind,][-12] #getting rid of "quality" variable
    df_currenttest <- df_current[-trainind,][-12]
    
    trainq <- as.vector(df_current[trainind,][12])$quality # "quality" column for training data
    testq <- as.vector(df_current[-trainind,][12])$quality # "quality" column for test data
    
    df_currentlogistic <- glm(highquality ~ ., data=df_currenttrain, family="binomial")

    response <- predict(df_currentlogistic, newdata = df_currenttest, type="response")
    response <- as.integer(response>0.5)
    
    tab <- table(Predicted = response, Actual = df_currenttest$highquality) #table of results
    errorrate[i] <- (tab[1,2]+tab[2,1])/sum(tab)
    for (i in 1:nrow(df_currenttest)){
        if ((df_currenttest$highquality)[i] != response[i]) {
            wrongs <- c(wrongs, testq[i])
        }
    }
  }
  results[[data_name]] = mean(errorrate)
  results_wrongs[[data_name]] = sum(wrongs==5 | wrongs==6)/length(wrongs)
}
results
results_wrongs
```

## Multiple Linear Regression
```{r, cache=TRUE}
results = list()
# as a reminder: datasets is a list with one item as white wine dataset and 
# another item as red wine dataset
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  for (i in 1:10){
    trainind <- sample(1:nrow(df_current), nrow(df_current)/2)
    train <- df_current[trainind,]
    test <- df_current[-trainind,]
    df_currentlm <- lm(quality ~ ., data=train)
    predictions <- predict(df_currentlm, newdata = test)
    
    mse[i] <- mean((predictions - test$quality)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

# Subset Selection
## Best Subset Selection
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    # best subset selection with max # of variables=11 on training data
    regfit.best=regsubsets(quality~.,data=df_current[train,],nvmax=11) 
    # create an X matrix of test data
    test.mat=model.matrix(quality~.,data=df_current[-train,]) 
    val.errors=rep(NA,11)
    for(j in 1:11){ 
       coefj=coef(regfit.best,id=j)
       pred=test.mat[,names(coefj)]%*%coefj
       # calculate MSE on validation set
       val.errors[j]=mean((df_current$quality[-train] - pred)^2)
    }
    # keep smallest MSE
    mse[i] = min(val.errors)
  }
  # average MSE
  results[[data_name]] = mean(mse)
}
results
```

## Forward Stepwise Selection
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    # forward selection with max # of variables=11 on training data
    regfit.forward=regsubsets(quality~.,data=df_current[train,],nvmax=11,
                           method="forward") 
    # create an X matrix of test data
    test.mat=model.matrix(quality~.,data=df_current[-train,]) 
    val.errors=rep(NA,11)
    for(j in 1:11){ 
       coefj=coef(regfit.forward,id=j)
       pred=test.mat[,names(coefj)]%*%coefj
       # calculate MSE on validation set
       val.errors[j]=mean((df_current$quality[-train] - pred)^2)
    }
    # keep smallest MSE
    mse[i] = min(val.errors)
  }
  # average MSE
  results[[data_name]] = mean(mse)
}
results
```

## Backward Stepwise Selection
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    # backward selection with max # of variables=11 on training data
    regfit.backward=regsubsets(quality~.,data=df_current[train,],nvmax=11,
                           method="backward") 
    # create an X matrix of test data
    test.mat=model.matrix(quality~.,data=df_current[-train,]) 
    val.errors=rep(NA,11)
    for(j in 1:11){ 
       coefj=coef(regfit.backward,id=j)
       pred=test.mat[,names(coefj)]%*%coefj
       # calculate MSE on validation set
       val.errors[j]=mean((df_current$quality[-train] - pred)^2)
    }
    # keep smallest MSE
    mse[i] = min(val.errors)
  }
  # average MSE
  results[[data_name]] = mean(mse)
}
results
```

# Shrinkage Methods
## Ridge Regression
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    x_train = model.matrix(quality~., df_current[train,])[,-1] # remove intercept
    y_train = df_current$quality[train]
    x_valid = model.matrix(quality~., df_current[-train,])[,-1] # remove intercept
    y_valid = df_current$quality[-train]
    # 10-fold CV to find best lambda
    cv.out=cv.glmnet(x_train,y_train,alpha=0,nfolds=10)
    bestlam=cv.out$lambda.min
    # training model
    ridge.mod=glmnet(x_train,y_train,alpha=0,lambda=bestlam)
    # predict on validation set
    pred = predict(ridge.mod, s=bestlam, newx=x_valid)
    mse[i] = mean((y_valid - pred)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

## Lasso Regression
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    x_train = model.matrix(quality~., df_current[train,])[,-1] # remove intercept
    y_train = df_current$quality[train]
    x_valid = model.matrix(quality~., df_current[-train,])[,-1] # remove intercept
    y_valid = df_current$quality[-train]
    # 10-fold CV to find best lambda
    cv.out=cv.glmnet(x_train,y_train,alpha=1,nfolds=10)
    bestlam=cv.out$lambda.min
    # training model
    lasso.mod=glmnet(x_train,y_train,alpha=1,lambda=bestlam)
    # predict on validation set
    pred = predict(lasso.mod, s=bestlam, newx=x_valid)
    mse[i] = mean((y_valid - pred)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

# Dimension Reduction Methods
## Principal Components Regression (PCR)
```{r, cache=TRUE}
results_pcr <- list()

for (data_name in names(datasets)) {
  df <- datasets[[data_name]]
  set.seed(5)
  mse_results <- rep(NA, 10)

  for (i in 1:10) {
    # Split data into training (50%) and validation (50%) sets
    train_index <- sample(1:nrow(df), nrow(df)/2)
    train_data <- df[train_index, ]
    test_data <- df[-train_index, ]

    # Max number of principal components to test
    max_comp <- min(ncol(train_data) - 1, nrow(train_data) - 1)

    # Cross-validation setup
    folds <- createFolds(train_data$quality, k = 10)
    cv_errors <- matrix(NA, nrow = 10, ncol = max_comp)

    for (fold in 1:10) {
      val_indices <- folds[[fold]]
      cv_train <- train_data[-val_indices, ]
      cv_val <- train_data[val_indices, ]

      # Fit PCR model since there's no auto cv do it manually
      pcr_model <- pcr(quality ~ ., data = cv_train, scale = TRUE, validation = "none")

      for (j in 1:max_comp) {
        pred <- predict(pcr_model, newdata = cv_val, ncomp = j)
        cv_errors[fold, j] <- mean((cv_val$quality - pred)^2)
      }
    }

    # Select optimal number of components from CV
    avg_cv_errors <- colMeans(cv_errors)
    best_comp <- which.min(avg_cv_errors)

    # Fit final PCR model with best number of components
    final_pcr_model <- pcr(quality ~ ., data = train_data, scale = TRUE, validation = "none")
    final_pred <- predict(final_pcr_model, newdata = test_data, ncomp = best_comp)

    mse_results[i] <- mean((test_data$quality - final_pred)^2)
  }

  results_pcr[[data_name]] <- mean(mse_results)  # Mean MSE 
}

results_pcr
```

## Partial Least Squares (PLS)
```{r, cache=TRUE}
results_pls <- list()

for (data_name in names(datasets)) {
  df <- datasets[[data_name]]
  set.seed(5)
  mse_results <- rep(NA, 10)

  for (i in 1:10) {
    # Split data into training (50%) and validation (50%) sets
    train_index <- sample(1:nrow(df), nrow(df)/2)
    train_data <- df[train_index, ]
    test_data <- df[-train_index, ]

    # Max number of PLS components to test
    max_comp <- min(ncol(train_data) - 1, nrow(train_data) - 1)

    # Cross-validation setup
    folds <- createFolds(train_data$quality, k = 10)
    cv_errors <- matrix(NA, nrow = 10, ncol = max_comp)

    for (fold in 1:10) {
      val_indices <- folds[[fold]]
      cv_train <- train_data[-val_indices, ]
      cv_val <- train_data[val_indices, ]

      # Fit PLS model (manual CV)
      pls_model <- plsr(quality ~ ., data = cv_train, scale = TRUE, validation = "none")

      for (j in 1:max_comp) {
        pred <- predict(pls_model, newdata = cv_val, ncomp = j)
        cv_errors[fold, j] <- mean((cv_val$quality - pred)^2)
      }
    }

    # Select optimal number of components from CV
    avg_cv_errors <- colMeans(cv_errors)
    best_comp <- which.min(avg_cv_errors)

    # Fit final PLS model with best number of components
    final_pls_model <- plsr(quality ~ ., data = train_data, scale = TRUE, validation = "none")
    final_pred <- predict(final_pls_model, newdata = test_data, ncomp = best_comp)

    mse_results[i] <- mean((test_data$quality - final_pred)^2)
  }

  results_pls[[data_name]] <- mean(mse_results)  # Mean MSE
}

results_pls
```

# Nonlinear Methods
## Polynomial Regression
```{r, cache = TRUE}
results_poly <- list()

for (data_name in names(datasets)) {
  df <- datasets[[data_name]]
  set.seed(5)
  mse_values <- rep(NA, 10)

  for (i in 1:10) {
    # Split data
    train_index <- sample(1:nrow(df), nrow(df)/2)
    train <- df[train_index, ]
    test <- df[-train_index, ]

    # Cross-validation to select best degree (1 to 10)
    cv_errors <- rep(NA, 4)
    for (d in 1:3) {
      predictors <- paste0("poly(", names(train)[-which(names(train) == "quality")],
                           ", ", d, ", raw = TRUE)", collapse = " + ")
      formula <- as.formula(paste0("quality ~ ", predictors))

      cv_model <- train(
        formula, data = train, method = "lm",
        trControl = trainControl(method="cv", number=10)
      )

      cv_errors[d] <- min(cv_model$results$RMSE)^2
    }

    best_degree <- which.min(cv_errors)

    # Fit best model on full training set
    final_predictors <- paste0("poly(", names(train)[-which(names(train) == "quality")], 
                               ", ", best_degree, ", raw = TRUE)", collapse = " + ")
    final_formula <- as.formula(paste0("quality ~ ", final_predictors))

    final_model <- lm(final_formula, data = train)
    predictions <- predict(final_model, newdata = test)
    mse_values[i] <- mean((predictions - test$quality)^2)
  }

  results_poly[[data_name]] <- mean(mse_values)  # Mean MSE
}

results_poly
```

## Basis Functions
```{r, cache = TRUE}
results_basis <- list()
poly_degrees <- 1:3

for (data_name in names(datasets)) {
  df <- datasets[[data_name]]
  set.seed(5)
  predictors <- df[, setdiff(names(df), "quality")]  # Keep all predictors

  mse <- matrix(NA, nrow=10, ncol=length(poly_degrees))

  for (i in 1:10) {
    train <- sample(1:nrow(df), nrow(df)/2)
    x_train <- predictors[train, ]
    y_train <- df$quality[train]
    x_valid <- predictors[-train, ]
    y_valid <- df$quality[-train]

    for (j in seq_along(poly_degrees)) {
      # Apply polynomial transformation to all predictors
      x_train_poly <- as.data.frame(lapply(x_train, 
                                           function(var) poly(var, poly_degrees[j])))
      x_valid_poly <- as.data.frame(lapply(x_valid, 
                                           function(var) poly(var, poly_degrees[j])))

      # Name transformed variables
      colnames(x_train_poly) <- unlist(lapply(names(x_train), 
                                              function(var) paste(var, "poly",
                                                                  1:poly_degrees[j],
                                                                  sep="_")))
      colnames(x_valid_poly) <- unlist(lapply(names(x_valid), 
                                              function(var) paste(var, "poly",
                                                                  1:poly_degrees[j],
                                                                  sep="_")))

      # Fit model using all transformed predictors
      poly_mod <- train(y_train ~ ., data=cbind(y_train, x_train_poly), method="lm",
                        trControl=trainControl(method="cv", number=10))

      # Predict and store MSE
      pred <- predict(poly_mod, newdata=x_valid_poly)
      mse[i, j] <- mean((y_valid - pred)^2)
    }
  }
  
  results_basis[[data_name]] <- mean(rowMeans(mse))  # Mean MSE across 10 iterations
}

results_basis
```

## Smoothing Splines
```{r, cache = TRUE}
results_smooth <- list()
df_range <- 2:10

for (data_name in names(datasets)) {
  df <- datasets[[data_name]]
  set.seed(5)
  predictors <- df[, setdiff(names(df), "quality")]
  mse_results <- rep(NA, 10)  # Store MSE results for validation iterations

  for (i in 1:10) {
    # Create 50/50 train-validation split
    train_index <- sample(1:nrow(df), nrow(df)/2)
    train_data <- df[train_index, ]
    valid_data <- df[-train_index, ]

    # Set up 10-fold CV
    folds <- createFolds(train_data$quality, k=10)
    cv_errors <- matrix(NA, nrow=10, ncol=length(df_range))

    for (fold in 1:10) {
      val_indices <- folds[[fold]]
      cv_train <- train_data[-val_indices, ]
      cv_val <- train_data[val_indices, ]

      for (j in seq_along(df_range)) {
        pred_matrix <- matrix(NA, nrow=nrow(cv_val), ncol=ncol(cv_train)-1)

        for (p in seq_along(names(cv_train)[-which(names(cv_train) == "quality")])) {
          predictor_name <- names(cv_train)[p]
          
          if (length(unique(cv_train[[predictor_name]])) > df_range[j]) {
            smooth_model <- smooth.spline(cv_train[[predictor_name]], cv_train$quality,
                                          df=df_range[j])
            pred_matrix[, p] <- predict(smooth_model, cv_val[[predictor_name]])$y
          } else {
            # Fallback for constant predictors
            pred_matrix[, p] <- mean(cv_train$quality, na.rm=TRUE)  
          }
        }

        cv_errors[fold, j] <- mean((cv_val$quality - rowMeans(pred_matrix))^2)
      }
    }

    # Find best df (minimum CV error)
    avg_cv_errors <- colMeans(cv_errors)
    best_df <- df_range[which.min(avg_cv_errors)]

    # Fit final model on full training set
    final_pred_matrix <- matrix(NA, nrow=nrow(valid_data), ncol=ncol(train_data)-1)
    for (p in seq_along(names(train_data)[-which(names(train_data) == "quality")])) {
      predictor_name <- names(train_data)[p]
      
      if (length(unique(train_data[[predictor_name]])) > best_df) {
        smooth_model <- smooth.spline(train_data[[predictor_name]], train_data$quality,
                                      df=best_df)
        final_pred_matrix[, p] <- predict(smooth_model, valid_data[[predictor_name]])$y
      } else {
        final_pred_matrix[, p] <- mean(train_data$quality, na.rm=TRUE)
      }
    }

    # Store MSE for this validation iteration
    mse_results[i] <- mean((valid_data$quality - rowMeans(final_pred_matrix))^2)
  }

  results_smooth[[data_name]] <- mean(mse_results)  # Mean MSE
}

results_smooth
```

## Generalized Additive Model (GAM)
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  for (i in 1:10){
    trainind <- sample(1:nrow(df_current), nrow(df_current)/2)
    
    df_currenttrain <- df_current[trainind,] #getting rid of quality col
    df_currenttest <- df_current[-trainind,]
    
    gam_model_df_current <- gam(quality ~ s(fixed.acidity) + 
                                  s(volatile.acidity) + s(citric.acid) + 
                                  s(residual.sugar) + s(chlorides) + 
                                  s(free.sulfur.dioxide) + 
                                  s(total.sulfur.dioxide) + s(density) + s(pH) +
                                  s(sulphates) + s(alcohol), 
                                data = df_currenttrain)
    
    pred_df_current <- predict(gam_model_df_current, df_currenttest)

    true_df_current <- df_currenttest$quality
    
    mse[i] <- mean((pred_df_current - true_df_current)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

# Tree-Based Methods
## Regression Tree 
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    # print(train)
    df_current.test = df_current[-train,"quality"]
    tree.df_current <- tree(quality~., data=df_current, subset=train)
    # choose best tree size based on k-fold CV lowest deviance
    cv.df_current = cv.tree(tree.df_current, K=10)
    best_size = cv.df_current$size[which.min(cv.df_current$dev)]
    # prune tree based on best tree size
    prune.df_current = prune.tree(tree.df_current, best=best_size)
    pred = predict(tree.df_current,newdata=df_current[-train,])
    mse[i] = mean((df_current.test - pred)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

## Bagging 
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    df_current.test = df_current[-train,"quality"]
    bag.df_current=randomForest(quality~.,data=df_current,subset=train,mtry=11)
    pred = predict(bag.df_current,newdata=df_current[-train,])
    mse[i] = mean((df_current.test - pred)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

## Random Forest 
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    df_current.test = df_current[-train,"quality"]
    rf.df_current=randomForest(quality~.,data=df_current,subset=train,mtry=3)
    pred = predict(rf.df_current,newdata=df_current[-train,])
    mse[i] = mean((df_current.test - pred)^2)
  }
  results[[data_name]] = mean(mse)
}
results
```

## Boosting 
```{r, cache=TRUE}
results = list()
for (data_name in names(datasets)) {
  df_current <- datasets[[data_name]]
  set.seed(5)
  mse = rep(NA,10)
  # validation set approach: repeat 10 times
  for (i in 1:10) {
    train = sample(1:nrow(df_current), nrow(df_current)/2)
    df_current.test=df_current[-train,"quality"]
    boost.df_current=gbm(quality~.,data=df_current[train,],distribution="gaussian")
    pred=predict(boost.df_current,newdata=df_current[-train,],n.trees=100)
    mse[i] = mean((df_current.test - pred)^2)
  }
  results[[data_name]] = mean(mse)
  }
results
```


